---
title: "Chapter 2: Statistical Learning"
author: "Sebastian Quirarte"
date: "2026-04-08"
output: 
  github_document: 
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Labs and exercises for the book 'An Introduction to Statistical Learning' by 
Gareth James, Daniela Witten, Trevor Hastie, and Rob Tibshirani. Full book is
available [here](https://www.statlearning.com/).

## Lab: Introduction to R

### Basic commands

```{r}
# Create a vector of numbers 
x <- c(1, 3, 2, 5)
x
```

```{r}
# Sum two vectors of the same length 
x <- c(1, 6, 2)
y <- c(1, 4, 3) 
x + y
```

```{r}
# View variables
ls()
# Delete variables 
rm(x, y)
ls()

# Delete all variables at once
rm(list = ls())
```

```{r}
# Create a matrix
x <- matrix(data = c(1, 2, 3, 4), nrow = 2, ncol = 2) # by column (default)
x

x <- matrix(data = c(1, 2, 3, 4), nrow = 2, ncol = 2, byrow = TRUE) # by row
x
```

```{r}
# sqrt and power can be applied to vectors and matrices
sqrt(x)
x^2
```

```{r}
# Random normal variables
x <- rnorm(50)
y <- x + rnorm(50, mean = 50, sd = 0.1)
# Correlation between variables
cor(x, y)
```

```{r}
# We can replicate results by setting a seed value
set.seed(1303)
rnorm(50)
```

```{r}
set.seed(3)
y <- rnorm(100)

mean(y) # mean 
var(y) # variance
sqrt(var(y)) # SD
sd(y) # SD
```

### Graphics

```{r}
x <- rnorm(100)
y <- rnorm(100)
plot(x, y)
plot(x, y, xlab = "This is the x-axis",
           ylab = "This is the y-axis",
           main = "Plot of X vs Y")
```

```{r}
# Sequence of numbers between a and b
x <- seq(1, 10)
x

x <- 1:10 # same as seq(1, 10)
x
```

```{r}
# Contour plot to present 3D data
y <- x
f <- outer(x, y, function(x, y) cos(y) / (1 + x^2))
contour(x, y, f)
contour(x, y, f, nlevels = 45, add = T)
fa <- (f - t(f)) / 2
contour(x, y, fa, nlevels = 15)

# image and persp can be used to view 3d plots
image(x, y, fa)
persp(x, y, fa)

# Theta and phi adjust the camara angle
persp(x, y, fa, theta = 30)
persp(x, y, fa, theta = 30, phi = 40)
```

### Indexing Data

```{r}
# Data stored in matrix
A <- matrix(1:16, 4, 4)
A

# Value at row 2, column 3
A[2, 3]

# Subsets of multiple rows and columns
A[c(1, 3), c(2, 4)]
A[1:3, 2:4]
A[1:2, ]
A[, 1:2]
A[1, ] # R treats a single row or column as a vector
A[-c(1,3), -c(1, 3, 4)] # Using a negative index indicates all except
dim(A) # dimensions of matrix (row, col)
```

### Loading Data

```{r}
# The Auto dataset is loaded from the ISLR2 package
library(ISLR2)
Auto <- Auto
head(Auto, header = T, na.string = '?')
dim(Auto) # rows x columns
names(Auto) # varible names
```

### Additional Graphical and Numerical Summaries

```{r}
plot(Auto$cylinders, Auto$mpg)
attach(Auto) # load variables locally so they can be called directly by name
plot(cylinders, mpg)

# Converting from quantitive to categorical
cylinders <- as.factor(cylinders)

# Plotting a qualitative variable on the x-axis will produce boxplots
plot(cylinders, mpg)

# Histograms can be plotted with the hist() function
hist(mpg)
hist(mpg, col = 2, breaks = 15)

# The pairs() function creates a scatterlot matrix
pairs(Auto)

plot(horsepower, mpg)

# The summary() function produces a summary numerical summary of variables in a dataset
summary(Auto)

# We can also provide a summary of a single variable
summary(mpg)
```

## Exercises 

### Conceptual

_**1. For each of parts (a) through (d), indicate whether we would generally
expect the performance of a flexible statistical learning method to be
better or worse than an inflexible method. Justify your answer.**_

*a) The sample size n is extremely large, and the number of predictors
p is small.*

*b) The number of predictors p is extremely large, and the number
of observations n is small.*

*c) The relationship between the predictors and response is highly
non-linear.*
Flexible method

*d) The variance of the error terms, i.e. σ2 = Var(ϵ), is extremely
high.*

2. Explain whether each scenario is a classification or regression problem,
and indicate whether we are most interested in inference or prediction.
Finally, provide n and p.



